\ch{RELIKE}: Reionization Effective Likelihood from Planck 2018 Data (?)

### 2021-04-09 paper/rei.tex

We publish RELIKE (Reionization Effective Likelihood), an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC). This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is being tested. We tested the code on two example models which showed good agreement with sampling the exact Planck likelihoods.
We also find that contrary to Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation. We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler. We expect the code release to dramatically reduce the time for testing individual \ch{global} ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as 21cm, luminosity function, star formation history, etc.
Finally, we derive model-independent constraints: $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and $\tau_{\rm PC}(15, 30) < 0.023 $ (95\% C.L.). The upper limits on high-redshift optical depth is at least a factor of 2 looser than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering  \ch{$\tau_{\rm 2step}(15, 30) < 0.013 $ (95\% C.L.)} using a  two-step toy model without using PCs.}

@@

Typically, the impact of reionization on the primary CMB fluctuations has been modeled as a steplike transition in the global ionization history \ch{in the form of a tanh function}, with the step location parameterized by the total Thomson optical depth induced. This steplike \ch{tanh} model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition. However, through the shape of the reionization bump induced in the CMB $E$-mode polarization at large angles, more information on the coarse-grained evolution of the ionization history can be obtained~\cite{Hu:2003gh, Mortonson:2007hq}.

@@

\ch{I'm here}
Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{−0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.

@@

\ch{write this part}
The code is publicly available on GitHub\footnote{[give link]}. [a bit more description here]. [comment on speed]. [comment on joint likelihood].

Extending PCs to cover up to $z_{\rm max} = 50$,we verified that there is no hint of ionization beyond $z>30$ in this Planck data release. But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is XX times less stringent than the Planck official results using Flex Knots \ch{[fill in number]} in Ref.~\cite{Aghanim:2018eyx}. Moreover, we demonstrate that the PC results of Ref.~\cite{Aghanim:2018eyx} suffers from a logical flaw in the treatment of its priors following Ref.
~\cite{Millea:2018bko} and ends up penalizing lower optical depth models. [quote difference in tau(15, 30) or total tau possibly attributed to this treatment]. %We follow the appendix of Ref.~\cite{Heinrich:2018btc} for the Planck 2015 data

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (red), Planck 2018 (blue, \texttt{SRoll2}, and Planck 2018 but with the \texttt{SimAll} likelihood (green). With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.  }
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

@@

Our code is available on GitHub at \todo{give URL\url{...}}. The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.  
This is also the default analysis for all results that follow except where noted.
The most important part of the code release is the python package called RELIKE which provides the PC parameters for a model and also
% added this since it now seemed to be emphasizing Gaussian only  - not sure that was intentional.
implements the Gaussian effective likelihood. The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use \textt{Relike} to return the effective likelihood of the model. The returned likelihood is already marginalized over cosmological parameters. 
To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample RELIKE within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}.

\wh{query: since this is the place where KDE is given not sure ``convenience" is the transition we want here.   Do you want to say something about KDE up front?}
we also provide a pre-assembled CosmoMC-Relike code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section. This CosmoMC-Relike code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.

@@

Instead of the KDE likelihood, we can alternatively use the Gaussian approximation given in Eq.~(\ref{eq:gaussian}). The Gaussian likelihood is faster to evaluate as it does not require looping over the entire $\sim 10^6$ points in the PC chain. \todo{[add timing here]} 
\wh{not sure we need timing results but if you do add them remember you can make Gaussian even faster by precomputing the linear combinations of $m_a$ that diagonalize the covariance (the first one should essentially be $\tau_{12}$ which is interesting pedagogically)} \ch{[CH: got it.]}
%
We obtain for the tanh model 
\beq
\tau_{\rm tanh} = 0.0592_{-0.0062}^{+0.0062}\;\; (\mathrm{Gaussian})
\eeq
Note that the Gaussian likelihood yields symmetric tanh $\tau$ constraints where as the KDE result 
is able to capture some of the skewness in the exact distribution. The shift in the peaks seen in Fig.~\ref{fig:tanh} between Gaussian and the exact likelihoods is consistent with this level of skewness ($\Delta \tau \sim 0.001$), which is $<0.2\sigma$. % 95\% UL:
%tanh Gauss: 1.996574506651048464e-03
%tanh KDE: 1.996656635642191091e-03
%tanh exact: 0.000000000000000000e+00

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{paper/plots/plot_xe_tanh_highz.pdf}
\caption{Two step ionization history $x_e(z)$ for the model of \refsec{example2}. In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$. Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L. $(\taulo, \tauhi) = (0.043, 0.026)$ (see also Fig.~\ref{fig:two_parameter_model_2D}). }
%best-fit (0.05290344, 0.005859431)
%95%CL (0.0427, 0.0256): zre = 6.4685, xemin = 0.0925 
\label{fig:two_step_model}
\end{figure}

@@

The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
 \bea
x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{1})-y(z)}{\Delta y} \right] \right\} \notag \\
&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
 \label{eqn:tanh_highz}
 \eea
where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z_1)^{1/2}\Delta z_1$ where $\Delta z_1 = 0.015\, (1+z_{1})$ just as in the first example for a sharper step.
We choose the second step at $z_{2}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
In comparison to tanh, a single parameter $\xemin$ is added in this model to control the high-$z$ ionization plateau for $z_{1} \lesssim z \lesssim z_2$. We would recover the standard tanh as $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$. We show examples
of this two-step model in Fig.~\ref{fig:two_step_model}.

@@

\todo{add timing results somewhere}

The marginalized 1D constraints from the KDE and Gaussian likelihoods 
\beq
\taulo = 0.0525^{+0.0062}_{-0.0079}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (KDE)}
\eeq 
and
\beq
\taulo = 0.0528^{+0.0064}_{- 0.0077}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (Gaussian)}
\eeq 
are consistent with those from the exact likelihood 
\beq
\taulo = 0.0526^{+0.0058}_{-0.0078}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (Exact).
\eeq 
}
%exact: tauhi < 0.0206
%KDE: tauhi < 0.0214
%Gaussian: tauhi < 0.0207

@@

Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$: \todo{give $\tau_{2\rm step}$.}

Finally the 95\% C.L. upper limit on high-redshift optical depth are also consistent:
\begin{equation}
    \tau_{2\text{step}}(15, 30) < 0.016 \quad (95\% \text{ C.L., Exact)}
    \label{eq:2stepbound}
\end{equation}
and 
$\tau_{2\text{step}}(15, 30) < 0.017$ for both effective likelihoods
Note also that the 95\% C.L. allowed two-step model in 
Fig.~\ref{fig:two_step_model} has
\ch{$\tau_{2\text{step}}(15,30)= 0.017$.}

@@

Using Eq.~(\ref{eq:cumtauPC}), we can derive constraints on the total optical
depth $\tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max})$ with the additional physicality priors on the PC amplitudes $m_a$ from
Eq.~\ref{eq:individualprior}.  These 
correspond to 
\beq
\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068} \;\; (\texttt{SRoll2}),
\label{eq:tau_pc_zmax30}
\eeq  
and the full posterior is shown in Fig.~\ref{fig:tauPC}. 
Notice that the distribution matches that of the two step model from \S \ref{sec:example2} but almost $0.5\sigma$ shifted higher than that of the tanh model from \S \ref{sec:example1}.

The similarity and difference between these results come down to implicit and explicit priors. As we have seen, 
results based on the tanh model requires that reionization happens suddenly at $z_{\rm re}$ and is maintained thereafter.   Therefore, the model has a strong implicit prior that a given a total optical depth $\tau$ comes from $z\lesssim z_{\rm re}$ regardless of whether the data allow or prefer
ionization at higher $z$.  While this prior may be well motivated
theoretically \cite{XX} (sims that give prompt reionization), it should be distinguished from constraints that are data driven.  
This was especially apparent with the Planck 2015 data where
the PC constraints gave \ch{$\tau=0.092\pm0.015$ which differed substantially from the tanh $\tau=0.079\pm0.017$~\cite{Heinrich:2016ojb}}.  As we shall discuss
below, these differences were even more striking in the implications for the cumulative $\tau$ at high redshift as we have also concretely illustrated in the two-step model of \S \ref{sec:example2}.

@@

Finally, we check that our results (Eq.~\ref{eq:tau_pc_zmax30}) are robust to
 extending the PC
range to
$z_{\rm max}=50$ which is negligibly different
\beq
\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{SRoll2}).
\eeq
Reverting the Planck 2018 likelihood to \texttt{SimAll} at $z_{\rm max}=30$ would give
\beq
\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{SimAll}).
\eeq
As with the tanh model, the \texttt{SRoll2} likelihood provides tighter constraints that shift the distribution toward higher optical depth.

@@

In Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior}, we show the 
the 68\% and 95\% confidence level contours for  $\tau_{\rm PC}(z,z_{\rm max})$ with and without the physicality prior on $m_1$.  Notice that without the prior, a larger range of unphysically negative optical depth is allowed whereas with the prior it is mostly eliminated except for the highest redshift.  Moreover upper bounds on the optical depth at high 
redshift are largely unchanged.   These upper limits allow some contribution at high-$z$ but do not require it.   For example
\beq
\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{SRoll2}).
\eeq
This should be contrasted with the tanh model 
 where it is  forbidden by form of the ionization history in conjunction with the total optical depth constraint.  
 On the other hand the bound is comparable to that obtained for the two-step model (see Eq.~\ref{eq:2stepbound}).  It is slightly larger given that the form of the two-step model still requires a steadily decreasing cumulative optical depth at $z>z_{\rm re}$).

@@

For the high-redshift optical depth, we also obtain consistent upper limits
%\beq
%\tau_{\rm PC}(15, 30)< 0.023 \; (95\%\; %\mathrm{C.L.})\;\;(\texttt{SRoll2}),
%\eeq
%[double check on full chain (from 10/15) this is from 09/30.]
%and 
\beq
\tau_{\rm PC}(15, 50) < 0.022 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{SRoll2}).
\eeq

\begin{figure}[ht]
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior_do_rescale_tau_false.pdf}
\caption{Cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ from the model-independent PC approach with and without the physicality prior.   The prior helps exclude models with negative contributions at high-$z$ but conservatively allows some negative values at $z \sim z_{\rm max}=30$ where the oscillations in the PCs no longer cancel due to the cumulative integration.  Upper limits are nearly unaffected.} 
}
% no prior: 95\% C.L.  = [-0.01935756790924691, 0.021029019067727237]
% with prior: {'z': 15, 'mean': 0.00694422689213187, 'std': 0.007568921831067864, 'limit68': [-0.0007215571045808687, 0.014684826477572082], 'limit95': [-0.006316156744039184, 0.02284538046347211]}
%
%}
\label{fig:plot_taugtz_2018_with_vs_without_physicality_prior}
\end{figure}

>>>>

We publish RELIKE (Reionization Effective Likelihood), an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC). This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is being tested. We tested the code on two example models which showed good agreement with sampling the exact Planck likelihoods.
We also find that contrary to Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation. We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler. We expect the code release to dramatically reduce the time for testing individual \ch{global} ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as 21cm, luminosity function, star formation history, etc.
Finally, we derive model-independent constraints: $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and $\tau_{\rm PC}(15, 30) < 0.023 $ (95\% C.L.). The upper limits on high-redshift optical depth is at least a factor of 2 looser than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering $\tau_{\rm 2step}(15, 30) < 0.017 $ (95\% C.L.) using a  two-step toy model without using PCs.}

@@

Typically, the impact of reionization on the primary CMB fluctuations has been modeled as a steplike transition in the global ionization history \ch{in the form of a tanh function}, with the step location parameterized by the total Thomson optical depth induced. This \ch{steplike tanh} model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition. However, through the shape of the reionization bump induced in the CMB $E$-mode polarization at large angles, more information on the coarse-grained evolution of the ionization history can be obtained~\cite{Hu:2003gh, Mortonson:2007hq}.

@@

\ch{I'm here}
Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{-0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.}

@@

\todo{write this part}
The code is publicly available on GitHub\footnote{[give link]}. [a bit more description here]. [comment on speed]. [comment on joint likelihood].

Extending PCs to cover up to $z_{\rm max} = 50$,we verified that there is no hint of ionization beyond $z>30$ in this Planck data release. But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is about \ch{$\sim 3$ times} less stringent than the Planck official results using Flex Knots in Ref.~\cite{Aghanim:2018eyx}. Moreover, we demonstrate that the PC results of Ref.~\cite{Aghanim:2018eyx} suffers from a logical flaw in the treatment of its priors following Ref.
~\cite{Millea:2018bko} and ends up penalizing lower optical depth models. [quote difference in tau(15, 30) or total tau possibly attributed to this treatment]. %We follow the appendix of Ref.~\cite{Heinrich:2018btc} for the Planck 2015 data

@@

\begin{figure}
\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (red), Planck 2018 (blue, \texttt{SRoll2}, and Planck 2018 but with the \texttt{SimAll} likelihood (green). With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.  \todo{add SRollv2. convention: put labels only when we do comparisons}}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

@@

Our code is available on GitHub at {give URL\url{...}}. The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.  
This is also the default analysis for all results that follow except where noted.
The most important part of the code release is the python package called RELIKE which provides the PC parameters for a model and also
% added this since it now seemed to be emphasizing Gaussian only  - not sure that was intentional.
implements the Gaussian effective likelihood. The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use \textt{Relike} to return the effective likelihood of the model. The returned likelihood is already marginalized over cosmological parameters. 
To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample RELIKE within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}.

\wh{query: since this is the place where KDE is given not sure ``convenience" is the transition we want here.   Do you want to say something about KDE up front?}
We also provide a pre-assembled CosmoMC code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section. This CosmoMC code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.

@@

Instead of the KDE likelihood, we can alternatively use the Gaussian approximation given in Eq.~(\ref{eq:gaussian}). The Gaussian likelihood is faster to evaluate as it does not require looping over the entire $\sim 10^6$ points in the PC chain. {[add timing here]} 
\wh{not sure we need timing results but if you do add them remember you can make Gaussian even faster by precomputing the linear combinations of $m_a$ that diagonalize the covariance (the first one should essentially be $\tau_{12}$ which is interesting pedagogically)} \ch{[CH: got it.]}
%
We obtain for the tanh model 
\beq
\tau_{\rm tanh} = 0.0592_{-0.0062}^{+0.0062}\;\; (\mathrm{Gaussian})
\eeq
Note that the Gaussian likelihood yields symmetric tanh $\tau$ constraints where as the KDE result 
is able to capture some of the skewness in the exact distribution. The shift in the peaks seen in Fig.~\ref{fig:tanh} between Gaussian and the exact likelihoods is consistent with this level of skewness ($\Delta \tau \sim 0.001$), which is $<0.2\sigma$. % 95\% UL:
%tanh Gauss: 1.996574506651048464e-03
%tanh KDE: 1.996656635642191091e-03
%tanh exact: 0.000000000000000000e+00

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{paper/plots/plot_xe_tanh_highz.pdf}
\caption{Two step ionization history $x_e(z)$ for the model of \refsec{example2}. In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$. Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L. $(\taulo, \tauhi) = (0.043, 0.026)$ (see also Fig.~\ref{fig:two_parameter_model_2D}). }\todo{CH: say in text 0.026 is 2D marginalized 95\%, 0.02 is 1D.}
%best-fit (0.05290344, 0.005859431)
%95%CL (0.0427, 0.0256): zre = 6.4685, xemin = 0.0925 
\label{fig:two_step_model}
\end{figure}

@@

\todo{change z1 z2 to zre and zhi everywhere.}
The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
 \bea
x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{1})-y(z)}{\Delta y} \right] \right\} \notag \\
&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
 \label{eqn:tanh_highz}
 \eea
where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z_1)^{1/2}\Delta z_1$ where $\Delta z_1 = 0.015\, (1+z_{1})$ just as in the first example for a sharper step.
We choose the second step at $z_{2}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
In comparison to tanh, a single parameter $\xemin$ is added in this model to control the high-$z$ ionization plateau for $z_{1} \lesssim z \lesssim z_2$. We would recover the standard tanh as $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$. We show examples
of this two-step model in Fig.~\ref{fig:two_step_model}.

@@

{add timing results somewhere}

The marginalized 1D constraints from the KDE and Gaussian likelihoods 
\beq
\taulo = 0.0525^{+0.0062}_{-0.0079}, \; \tauhi < 0.021 \quad (95\% \text{ C.L., KDE)}
\eeq 
and
\beq
\taulo = 0.0528^{+0.0064}_{- 0.0077}, \; \tauhi < 0.021 \quad (95\% \text{ C.L., Gaussian)}
\eeq 
are consistent with those from the exact likelihood 
\beq
\taulo = 0.0526^{+0.0058}_{-0.0078}, \; \tauhi < 0.021 \quad (95\% \text{ C.L., Exact).
\eeq 
}
%exact: tauhi < 0.0206
%KDE: tauhi < 0.0214
%Gaussian: tauhi < 0.0207

@@

\ch{comment here about 2D 95\% model picked with tau hi = 0.026 is higher because 0.021 is the 1D marginalized result.}
Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$:
\ch{
\beq
\tau_{\rm 2step} = 0.0621^{+0.0062}_{-0.0061}.
\eeq
}
Finally the 95\% C.L. upper limit on high-redshift optical depth are also consistent:
\begin{equation}
    \tau_{2\text{step}}(15, 30) < 0.016 \quad (95\% \text{ C.L., Exact)}
    \label{eq:2stepbound}
\end{equation}
and 
$\tau_{2\text{step}}(15, 30) < 0.017$ for both effective likelihoods
Note also that the 95\% C.L. allowed two-step model in 
Fig.~\ref{fig:two_step_model} has
\ch{$\tau_{2\text{step}}(15,30)= 0.017$.}\todo{Need to change all 95\% upper limit on tau(15,30) to a one-sided computation instead of 2.}\todo{update plot two-step with longer chain results, run 2 more days exact two-step chains.} \todo{point out in text that 95\% upper limit on tau(15,30) not the same as plot.}

@@

Using Eq.~(\ref{eq:cumtauPC}), we can derive constraints on the total optical
depth $\tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max})$ with the additional physicality priors on the PC amplitudes $m_a$ from
Eq.~\ref{eq:individualprior}.  These 
correspond to 
%
\beq
\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068},
\label{eq:tau_pc_zmax30}
\eeq  
%
and the full posterior is shown in Fig.~\ref{fig:tauPC}. 
Notice that the distribution matches that of the two step model from \S \ref{sec:example2} but almost $0.5\sigma$ shifted higher than that of the tanh model from \S \ref{sec:example1}.

The similarity and difference between these results come down to implicit and explicit priors. As we have seen, 
results based on the tanh model requires that reionization happens suddenly at $z_{\rm re}$ and is maintained thereafter.   Therefore, the model has a strong implicit prior that a given a total optical depth $\tau$ comes from $z\lesssim z_{\rm re}$ regardless of whether the data allow or prefer
ionization at higher $z$.  While this prior may be well motivated
theoretically \cite{XX} (sims that give prompt reionization), it should be distinguished from constraints that are data driven.  
This was especially apparent with the Planck 2015 data where
the PC constraints gave \ch{$\tau_{\rm PC}=0.092\pm0.015$ which differed substantially from the tanh $\tau_{\rm PC}=0.079\pm0.017$~\cite{Heinrich:2016ojb}}.  As we shall discuss
below, these differences were even more striking in the implications for the cumulative $\tau$ at high redshift as we have also concretely illustrated in the two-step model of \S \ref{sec:example2}.

@@

Finally, we check that our results (Eq.~\ref{eq:tau_pc_zmax30}) are robust to
 extending the PC
range to
$z_{\rm max}=50$ which is negligibly different
\ch{
\beq
%\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{SRoll2}).
\tau_{\rm PC} = 0.0626^{+0.0061}_{-0.0072}\quad (\zmax=50).
\eeq
CH: removing SRoll2 label in parenthesis and moving zmax=50 into the parenthesis.}
Reverting the Planck 2018 likelihood to \texttt{SimAll} at $z_{\rm max}=30$ would give
\beq
\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{SimAll}).
\eeq
As with the tanh model, the \texttt{SRoll2} likelihood provides tighter constraints that shift the distribution toward higher optical depth.

@@

In Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior}, we show the 
the 68\% and 95\% confidence level contours for  $\tau_{\rm PC}(z,z_{\rm max})$ with and without the physicality prior on $m_1$.  Notice that without the prior, a larger range of unphysically negative optical depth is allowed whereas with the prior it is mostly eliminated except for the highest redshift.  Moreover upper bounds on the optical depth at high 
redshift are largely unchanged.   These upper limits allow some contribution at high-$z$ but do not require it.   For example,
\ch{
\beq
\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.}).
\eeq
}
This should be contrasted with the tanh model 
 where it is  forbidden by form of the ionization history in conjunction with the total optical depth constraint.  
 On the other hand the bound is comparable to that obtained for the two-step model (see Eq.~\ref{eq:2stepbound}).  It is slightly larger given that the form of the two-step model still requires a steadily decreasing cumulative optical depth at $z>z_{\rm re}$).
 \wh{I might add a few words about the relationship to $\tau_{\rm hi}$ but first let's make sure that all upper limits are quoted as one sided 95\% since I'm not sure we're comparing apples-to-apples yet..}

@@

For the high-redshift optical depth, we also obtain consistent upper limits
%\beq
%\tau_{\rm PC}(15, 30)< 0.023 \; (95\%\; %\mathrm{C.L.})\;\;(\texttt{SRoll2}),
%\eeq
%[double check on full chain (from 10/15) this is from 09/30.]
%and 
\beq
\tau_{\rm PC}(15, 50) < 0.022 \; (95\%\; \mathrm{C.L.}).
\eeq

\begin{figure}[ht]
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior_do_rescale_tau_false.pdf}
\caption{Cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ from the model-independent PC approach with and without the physicality prior.   The prior helps exclude models with negative contributions at high-$z$ but conservatively allows some negative values at $z \sim z_{\rm max}=30$ where the oscillations in the PCs no longer cancel due to the cumulative integration.  Upper limits are nearly unaffected. \todo{change all tau cumulative plots for PC to $\tau_{\rm PC}$;} \todo{add (with prior)}} 
}
% no prior: 95\% C.L.  = [-0.01935756790924691, 0.021029019067727237]
% with prior: {'z': 15, 'mean': 0.00694422689213187, 'std': 0.007568921831067864, 'limit68': [-0.0007215571045808687, 0.014684826477572082], 'limit95': [-0.006316156744039184, 0.02284538046347211]}
%
%}
\label{fig:plot_taugtz_2018_with_vs_without_physicality_prior}
\end{figure}
